State-space analysis of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms

### 2023-11-27 manuscript/manuscript_2/paper.tex
Eod
n4000, low SNR

 \textcolor{red}{TK: there is an obvious additional feature that I haven't yet mentioned in the text, which is the systematic offset in iota and h w.r.t the injection. The systematic-ness makes me think this is likely a sampler setting. I am currently running some checks, and will then add few sentences discussing this, or update the results/plots.}

@@

@@

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR}
	\caption{Same as Figure \ref{fig:corner_plot_compare_high}, but for a low-SNR system with $h_0 = 5 \times 10^{-15}$. The posterior distributions from the Earth-term and pulsar-term models are generally overlaid }
	\label{fig:corner_plot_compare_low}
\end{figure*}

@@

\section{Kalman filter} \label{sec:kalman}
In this appendix, we describe the Kalman filter algorithm used in this paper. General recursion relations for the discrete-time Kalman filter are written down for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}. The mapping onto the specific continuous-time state-space model in Section \ref{sec:model} is written down in Section \ref{sec_kalman_specific}.

The Kalman filter \citep{Kalman1} is a Gauss-Markov model used to algorithmically recover a temporal sequence of stochastically evolving  system state variables, $\boldsymbol{X}(t)$, which are not observed directly, given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$. It finds common use in engineering applications and has been applied successfully in neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}. In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement), with ${\boldsymbol{X}}(t) = \{ f_{\rm p}^{(n)}(t) \}$ and ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$. Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}. Equations \eqref{eq:measurement} and \eqref{eq:g_func} are non-linear in the static parameters (e.g.\ $d^{(n)}$), even though they are linear in ${\boldsymbol{X}}(t)$ and ${\boldsymbol{Y}}(t)$. Hence inferring the static parameters is a non-linear exercise, to be tackled separately after the linear Kalman filter operates on the data for a fixed set of static parameters. Inference of the static parameters in Equations \eqref{eq:measurement} and \eqref{eq:g_func} by nested sampling is discussed in Section \ref{sec:nested_sampling}. \newline

Implementation of the linear Kalman filter for the PTA state-space model in Section \ref{sec:model}, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}. At each discrete timestep indexed by $ 1 \leq i  \leq M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i = \hat{\boldsymbol{X}}(t_i)$, and the covariance of those estimates, ${\boldsymbol{P}}_i = \langle {\boldsymbol{\hat X}}_i {\boldsymbol{\hat X}}_i^{\rm T} \rangle$, where the superscript T denotes the matrix transpose. The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by converting ${\boldsymbol{\hat X}}_i$ into predicted measurements ${\boldsymbol{\hat Y}}_i$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func} and comparing with the actual, noisy measurements ${\boldsymbol{\hat Y}}_i$. This defines a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, which is sometimes termed the innovation. The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{T} \rangle$. The innovation and the innovation covariance are then used to correct the state estimate ${\boldsymbol{\hat X}}_i$ according to Equation \eqref{eq:kalmangainupdate}. For a fixed set of static parameters, the Kalman filter returns an estimate of the state sequence ${\boldsymbol {\hat X}}_1, \dots , {\boldsymbol{\hat X}}_M$ which minimizes the mean square error. We explain how to use this intermediate output to infer the optimal values of the static parameters ${\boldsymbol{\theta}}$ in Section \ref{sec:nested_sampling} \newline

The Gaussian log-likelihood of obtaining ${\boldsymbol{Y}}_i$ given ${\boldsymbol{\hat X}}_i$ can  then be calculated at each timestep from the Kalman filter output according to
\begin{eqnarray}
	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ .
\end{eqnarray}
The total log-likelihood for the entire sequence is
\begin{eqnarray}
	\log \mathcal{L} =  \sum_{i=1}^{M} \log \mathcal{L}_i \ . \label{eq:likelihood}
\end{eqnarray}
Given ${\boldsymbol{Y}}_1, \dots, {\boldsymbol{Y}}_M$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e. $\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\hat \theta})$. Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of $\boldsymbol{\hat \theta}$. If $\boldsymbol{\hat{\theta}}$ is close to the true underlying parameters $\boldsymbol{\theta}$, then the errors in $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$ are minimized and $\mathcal{L}$ is maximised. This is illustrated with synthetic data in Figure \ref{fig:kalman_example}. In the left column, a time series of $f_{\rm m}^{(1)}(t)$ including Gaussian noise (middle panel, red curve) is generated from Equations \eqref{eq:spinevol}-- \eqref{eq:xieqn}, \eqref{eq:measurement}, and \eqref{eq:g_func} for a single pulsar and fed into the Kalman filter along with the true static parameters ${\boldsymbol{\hat \theta}} = {\boldsymbol{\theta}}$. The Kalman filter recovers the evolution of $f_{\rm p}^{(1)}(t)$ with high fidelity; the estimate of $\hat{f}_{\rm p}^{(1)}(t)$ (left top panel, blue curve) overlaps almost perfectly with the true $f_{\rm p}^{(1)}(t)$ (left top panel, green curve). The predicted state $\hat{f}_{\rm p}^{(1)}(t)$ is converted  into a predicted measurement $\hat{f}_{\rm m}^{(1)}(t)$ (middle panel, magenta curve), which again overlaps almost perfectly with the true measurement. The residuals $\epsilon(t)$ between the true and predicted measurements are small ($\lesssim 0.1\%$) and normally distributed (left bottom panel). By contrast, in the right column, the exercise is repeated while passing slightly incorrect static parameters (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$) to the Kalman filter, where $\Omega$ is displaced from its true value by $20 \%$. In this case the Kalman filter fails to track $f_{\rm p}^{(1)}(t)$ accurately, as the discrepancy between the blue and green curves in the top panel of the right-hand column indicates. It similarly fails to predict $f_{\rm m}^{(1)}(t)$ accurately, as shown by the discrepancy between the red and magenta curves in the middle panel, and the residuals are no longer distributed normally (right bottom panel). In Section \ref{sec:nested_sampling}, we explain how to combine the Kalman filter with a nested sampler to iteratively guide ${\boldsymbol{\hat \theta}}$ towards the true value of ${\boldsymbol{\theta}}$.

\subsection{Recursion equations}\label{sec_kalman_general}
The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation
\begin{equation}
	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
\end{equation}
where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step. The Kalman filter evolves the underlying states according to
\begin{equation}
	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
\end{equation}
where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ is the control matrix. $\boldsymbol{u}_k$ is the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$ with covariance $\boldsymbol{Q}_k$ \newline

The Kalman filter is a recursive estimator with two distinct stages: a ``predict" stage and an ``update" stage. The predict stage predicts $\hat{\boldsymbol{X}}_{k|k-1}$, the estimate of the state at discrete step $k$, given the state estimates from step $k-1$. Specifically, the predict step proceeds as
\begin{align}
	\hat{\boldsymbol{X}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k \ , \\
		\hat{\boldsymbol{P}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k  \ ,
\end{align}
where $\hat{\boldsymbol{P}}_{k|k-1}$ is the covariance of the prediction. Note that the predict stage is independent of the measurements. The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
\begin{align}
	\boldsymbol{\epsilon}_{k} &= \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1} \ , \\
	\boldsymbol{S}_k &= \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k \ , \\
	\boldsymbol{K}_k &= \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \ ,\label{eq:kalman gain} \\
		\hat{\boldsymbol{X}}_{k|k} &=\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \ , \label{eq:kalmangainupdate} \\
			\hat{\boldsymbol{P}}_{k|k} &= \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1} \ .
\end{align}
Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.  $\boldsymbol{K}_k = \text{argmin} \left \{ \boldsymbol{E}[ (\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k)^2 ] \right \}$. For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}. \newline

To apply the Kalman filter in practice means specifying the eight component matrices that make up the ``machinery'' of the filter: $\boldsymbol{X}_k$, $\boldsymbol{Y}_k$, $\boldsymbol{F}_k$, $\boldsymbol{G}_k$, $\boldsymbol{u}_k$, $\boldsymbol{H}_k$, $\boldsymbol{Q}_k$ and $\boldsymbol{R}_k$. In Section \ref{sec_kalman_specific} we describe how the machinery is determined for the state-space model in Section \ref{sec:model}.

\subsection{State space representation of a PTA analysis}\label{sec_kalman_specific}
We apply the Kalman recursion relations in Section \ref{sec_kalman_general} to the state-space model of a PTA with $N$ pulsars described in Section \ref{sec:model} as follows. \newline

@@

In the idealized model above, each pulsar's rotational state evolves phenomenologically according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, whose general solution is given by \citep{gardiner2009stochastic},
\begin{equation}
	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \ . \label{eq:gardenier}
\end{equation} 
From Equation \eqref{eq:gardenier} we construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2}, with
\begin{align}
	\boldsymbol{F}_k &= e^{\boldsymbol{A} \Delta t } \  \\
	&= \text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right) \ ,
\end{align}
\begin{align}
	\boldsymbol{G}_k \boldsymbol{u}_k &= \int_{t_k}^{t_{k+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' \ , \\
	&= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right) ,
\end{align}
\begin{equation}
	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t') \ ,  \label{eq:appendix_noise}
\end{equation}
\begin{align}
	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left(\Delta t + t_k \right) \nonumber \\ 
	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right] \ ,
\end{align}
and $\Delta t = t_{k+1} - t_k$. From Equation \eqref{eq:appendix_noise} the process noise covariance matrix is
\begin{equation}
	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) \ ,
\end{equation}
with 
\begin{equation}
Q^{(n)} = \frac{[\sigma^{n}]^2}{2 \gamma^{(n)}} \left[ 1 - e^{-2 \gamma^{(n)} \Delta t}\right] \ .
\end{equation}

The two remaining unspecified component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$. These are defined straightforwardly from Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}. Specifically, 
$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}. The measurement covariance satisfies $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$ for all $k$.

\section{Nested sampling}\label{sec:nested_sampling}

We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper. Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}. It also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$. It does so by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point. The primary advantage of nested sampling is its ability to compute $\mathcal{Z}$, on which model selection relies. Nested sampling is also computationally efficient and can handle multi-modal problems \citep{Ashton2022}. For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, neutron star astrophysics \citep{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}. For reviews of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}. Multiple nested sampling algorithms and computational libraries exist. \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}. In gravitational wave research it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library. We follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work. \newline

>>>>

 \textcolor{red}{TK: there is an obvious additional feature that I haven't yet mentioned in the text, which is the systematic offset in iota and h w.r.t the injection. The systematic-ness makes me think this is likely a sampler setting. I am currently running some checks, and will then add few sentences discussing this, or update the results/plots. This is not an issue in the high SNR case: can noise drive results away from the injection systematically? Naively I expect a scatter, symmetric about the injection... }

@@

With a strain of X, we are close to the nosie floor (see e.g. figure X), so small variations in

@@

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR_compare}
	\caption{Same as Figure \ref{fig:corner_plot_compare_high}, but for a low-SNR system with $h_0 = 5 \times 10^{-15}$. The posterior distributions from the Earth-term and pulsar-term models are generally overlaid }
	\label{fig:corner_plot_compare_low}
\end{figure*}

@@

\section{Kalman filter} \label{sec:kalman}
The Kalman filter \citep{Kalman1} is a Gauss-Markov model used to algorithmically recover a temporal sequence of stochastically evolving  system state variables, $\boldsymbol{X}(t)$, which are not observed directly, given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$. It finds common use in engineering applications and has been applied successfully in neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}. In this appendix, we describe the Kalman filter algorithm used in this paper. General recursion relations for the discrete-time Kalman filter are written down for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}, along with the calculation of the likelihood. The application of the Kalman filter algorithm to the specific continuous-time state-space model in Section \ref{sec:2} is outlined in Section \ref{sec_kalman_specific}.

%Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, unscented Kalman filter \citep{882463van} or particle filter \citep{Simon10}. Equations \eqref{eq:measurement} and \eqref{eq:g_func} are non-linear in the static parameters (e.g.\ $d^{(n)}$), even though they are linear in ${\boldsymbol{X}}(t)$ and ${\boldsymbol{Y}}(t)$.
 %Implementation of the linear Kalman filter for the PTA state-space model in Section \ref{sec:model}, including the full set of Kalman recursion relations is presented in Appendix \ref{sec:kalman}.
%
%At each discrete timestep indexed by , the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i = \hat{\boldsymbol{X}}(t_i)$, and the covariance of those estimates, ${\boldsymbol{P}}_i = \langle {\boldsymbol{\hat X}}_i {\boldsymbol{\hat X}}_i^{\rm T} \rangle$, where the superscript T denotes the matrix transpose.

\subsection{Recursion equations and likelihoood }\label{sec_kalman_general}
In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement), with ${\boldsymbol{X}}(t) = \{ f_{\rm p}^{(n)}(t) \}$ and ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$. The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k = \boldsymbol{Y}(t_k)$, which are related via a linear transformation to a set of unobservable discrete system states $\boldsymbol{X}_k= \boldsymbol{Y}(t_k)$. Each discrete timestep is indexed by $ 1 \leq k  \leq K$. The measurements are related to the states via
\begin{equation}
	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
\end{equation}
where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step. The Kalman filter evolves the underlying states according to
\begin{equation}
	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
\end{equation}
where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ is the control matrix. $\boldsymbol{u}_k$ is the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$ with covariance $\boldsymbol{Q}_k$ \newline

The Kalman filter is a recursive estimator with two distinct stages: a ``predict" stage and an ``update" stage. The predict stage predicts $\hat{\boldsymbol{X}}_{k|k-1}$, the estimate of the state at discrete step $k$, given the state estimates from step $k-1$. Specifically, the predict step proceeds as
\begin{align}
	\hat{\boldsymbol{X}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k \ , \\
	\hat{\boldsymbol{P}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k  \ ,
\end{align}
where $\hat{\boldsymbol{P}}_{k|k-1}$ is the covariance of the prediction. Note that the predict stage is independent of the measurements. The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
\begin{align}
	\boldsymbol{\epsilon}_{k} &= \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1} \ , \label{eq:residual} \\
	\boldsymbol{S}_k &= \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k \ , \label{eq:innv_covar}\\
	\boldsymbol{K}_k &= \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \ ,\label{eq:kalman gain} \\
	\hat{\boldsymbol{X}}_{k|k} &=\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \ , \label{eq:kalmangainupdate} \\
	\hat{\boldsymbol{P}}_{k|k} &= \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1} \ .
\end{align}
Equation \eqref{eq:residual} defines a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, which is sometimes termed the innovation. The uncertainty in $\boldsymbol{\epsilon}_i$ is quantified via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{T} \rangle$, Equation \eqref{q:innv_covar}. Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.  $\boldsymbol{K}_k = \text{argmin} \left \{ \boldsymbol{E}[ (\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k)^2 ] \right \}$. For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}. \newline

The Gaussian log-likelihood of obtaining ${\boldsymbol{Y}}_i$ given ${\boldsymbol{\hat X}}_i$ is calculated at each timestep from the Kalman filter output according to
\begin{eqnarray}
	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ ,
\end{eqnarray}
where $N$ is the number of data points at timestep $i$. The total log-likelihood for the entire sequence is
\begin{eqnarray}
	\log \mathcal{L} =  \sum_{i=1}^{M} \log \mathcal{L}_i \ . \label{eq:likelihood}
\end{eqnarray}
Given ${\boldsymbol{Y}}_1, \dots, {\boldsymbol{Y}}_M$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e. $\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\hat \theta})$. Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of $\boldsymbol{\hat \theta}$. In Section \ref{sec:nested_sampling}, we explain how to combine the Kalman filter with a nested sampler to iteratively guide ${\boldsymbol{\hat \theta}}$ towards the true value of ${\boldsymbol{\theta}}$. \newline

To apply the Kalman filter in practice means specifying the eight component matrices that make up the ``machinery'' of the filter: $\boldsymbol{X}_k$, $\boldsymbol{Y}_k$, $\boldsymbol{F}_k$, $\boldsymbol{G}_k$, $\boldsymbol{u}_k$, $\boldsymbol{H}_k$, $\boldsymbol{Q}_k$ and $\boldsymbol{R}_k$. In Section \ref{sec_kalman_specific} we describe how the machinery is determined for the state-space model in Section \ref{sec:2}.

\subsection{State space representation of a PTA analysis}\label{sec_kalman_specific}
We apply the Kalman recursion relations in Section \ref{sec_kalman_general} to the state-space model of a PTA with $N$ pulsars described in Section \ref{sec:2} as follows. \newline

@@

In the idealized model above, each pulsar's rotational state evolves phenomenologically according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, whose general solution is given by \citep{gardiner2009stochastic},
\begin{equation}
	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \ . \label{eq:gardenier}
\end{equation} 
From Equation \eqref{eq:gardenier} we construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2}, with
\begin{align}
	\boldsymbol{F}_k &= e^{\boldsymbol{A} \Delta t } \  \\
	&= \text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right) \ ,
\end{align}
\begin{align}
	\boldsymbol{G}_k \boldsymbol{u}_k &= \int_{t_k}^{t_{k+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' \ , \\
	&= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right) ,
\end{align}
\begin{equation}
	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t') \ ,  \label{eq:appendix_noise}
\end{equation}
\begin{align}
	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left(\Delta t + t_k \right) \nonumber \\ 
	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right] \ ,
\end{align}
and $\Delta t = t_{k+1} - t_k$. From Equation \eqref{eq:appendix_noise} the process noise covariance matrix is
\begin{equation}
	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) \ ,
\end{equation}
with 
\begin{equation}
Q^{(n)} = \frac{[\sigma^{n}]^2}{2 \gamma^{(n)}} \left[ 1 - e^{-2 \gamma^{(n)} \Delta t}\right] \ .
\end{equation}
The two remaining unspecified component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$. These are defined straightforwardly from Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}. Specifically, 
$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func_trig}. The measurement covariance satisfies $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$ for all $k$.

\section{Nested sampling}\label{sec:nested_sampling}
We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper. Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}. It also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$. It does so by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point. The primary advantage of nested sampling is its ability to compute $\mathcal{Z}$, on which model selection relies. Nested sampling is also computationally efficient and can handle multi-modal problems \citep{Ashton2022}. For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, neutron star astrophysics \citep{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}. For reviews of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}. Multiple nested sampling algorithms and computational libraries exist. \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}. In gravitational wave research it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library. We follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work. \newline
